### 2015-06-06 chapters/ipa_meets_unicode.tex
started working on pitfall IPA homoglyphs

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
use the same glyph. If this distinction has to be kept upright in italics, the
only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ>. However, note that when the text is transferred to another font
(i.e.~one that includes real italics) and/or to another software application
(like Apple Pages, which does not perform slanting), then this visual appearance
will be lost. In this case we are thus still in the pre-Unicode situation in
which the choice of font and rendering software actually matters. The ideal
solution from a linguistic point of view would be the introduction of a new IPA
code point for a different kind of which explicitly specifies that it should
still be rendered as a double-story character when italicized. After informal
discussion with various Unicode players, our impression is that this highly
restricted problem is not sufficiently urgent to introduce even more <a>-like
characters in Unicode (which already lead to much confusion, see Section~
\ref{pitfall-homoglyphs}).}

@@

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode
       Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
       at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
       \textsc{right single quotation mark} at \uni{2019}.
     \item The glottal stop/glottalization/ejective marker is another completely
        different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
        \uni{02BC}, but unfortunately looks mostly highly similar to
        \uni{2019}. 
	\item Another problem is the <ˁ> \textsc{modifier letter reversed
       glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
       letter small reversed glottal stop} at \uni{02E4}. Both are denoted in
       the Unicode Standard as the \textsc{pharyngealized diacritic} and both
       appear in various resources representing phonetic data online. This is
       thus an example for which the Unicode Standard does not solve the
       linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the
       Unicode Consortium does not match the IPA's description. In the Unicode
       Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
       click}, but in IPA is an alveolar or postalveolar click (not retroflex).
       This naming is probably best seen as a simple error in the Unicode
       Standard. This character is of course often simply typed as <!>
       \textsc{exclamation mark} at \uni{0021}.
\end{itemize}

@@

It is not just the Unicode Standard that offers multiple options for encoding 
the IPA. Even the IPA itself is not consistent in how specific meaningful 
transcriptions have to be encoded. There are a few cases in which the IPA 
explicitly allows for different options of transcribing the same phonetic 
content

Another example
we commonly encounter is the use of <g> \textsc{latin small letter g} at \uni{0067},
instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to question
whether this issue is at all apparent to the working linguist, or if they simply
use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
the latter must be cumbersomely inserted as a special symbol in most
software.

\begin{itemize}
  \item tone marks (prefer tone symbols)
  \item marking of <g>\footnote{This issue was recently addressed by the International
Phonetic Association, which has taken the stance that both the keyboard
\textsc{latin small letter g} and the \textsc{latin small letter script g} are
valid input characters for the voiced velar plosive. Unfortunately, this
decision further introduces ambiguity for linguists trying to adhere to a strict
Unicode Standard IPA encoding.}
  \item marking of voiceless with ring diacritic
  \item tie bar
\end{itemize}

>>>>

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
use the same glyph. If this distinction has to be kept upright in italics, the
only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ>. However, note that when the text is transferred to another font
(i.e.~one that includes real italics) and/or to another software application
(like Apple Pages, which does not perform slanting), then this visual appearance
will be lost. In this case we are thus still in the pre-Unicode situation in
which the choice of font and rendering software actually matters. The ideal
solution from a linguistic point of view would be the introduction of a new IPA
code point for a different kind of which explicitly specifies that it should
still be rendered as a double-story character when italicized. After informal
discussion with various Unicode players, our impression is that this highly
restricted problem is not sufficiently urgent to introduce even more <a>-like
characters in Unicode (which already lead to much confusion, see Section~\ref{pitfall-homoglyphs}).}

@@

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode
       Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
       at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
       \textsc{right single quotation mark} at \uni{2019}.
     \item The glottal stop/glottalization/ejective marker is another completely
        different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
        \uni{02BC}, but unfortunately looks mostly highly similar to
        \uni{2019}. 
	\item Another problem is the <ˁ> \textsc{modifier letter reversed
       glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
       letter small reversed glottal stop} at \uni{02E4}. Both 
       appear in various resources representing phonetic data online. This is
       thus a clear example for which the Unicode Standard does not solve the
       linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the
       Unicode Consortium does not match the IPA's description. In the Unicode
       Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
       click}, but in IPA that glyph is used for an alveolar or postalveolar
       click (not retroflex). This naming is probably best seen as a simple
       error in the Unicode Standard. This character is of course often simply
       typed as <!> \textsc{exclamation mark} at \uni{0021}. \end{itemize}

@@

It is not just the Unicode Standard that offers multiple options for encoding 
the IPA. Even the IPA itself offers some flexibility in how specific meaningful 
transcriptions have to be encoded. There are a few cases in which the IPA 
explicitly allows for different options of transcribing the same phonetic 
content. This is understandable from a transcriber's point of view, but it is 
not good interoperability between resources supposedly written in IPA. We 
consider it crucial to distinguish between ``lax'' IPA, in which the 
phonetically-trained reader should be able to understand the transcription, and 
``strict'' IPA, which should be standardized on a single encoding so search will 
work across resources.

Another example we commonly encounter is the use of <g> \textsc{latin small
letter g} at \uni{0067}, instead of the Unicode Standard IPA character for the
voiced velar stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One
begins to question whether this issue is at all apparent to the working
linguist, or if they simply use the \uni{0067} because it is easily keyboarded
and thus saves time, whereas the latter must be cumbersomely inserted as a
special symbol in most software.\footnote{This issue was recently addressed by the International
Phonetic Association, which has taken the stance that both the keyboard
\textsc{latin small letter g} and the \textsc{latin small letter script g} are
valid input characters for the voiced velar plosive. Unfortunately, this
decision further introduces ambiguity for linguists trying to adhere to a strict
Unicode Standard IPA encoding.}

\begin{itemize}
  \item tone marks (prefer tone symbols)
  \item marking of <g>  \item marking of voiceless with ring diacritic
  \item tie bar
\end{itemize}
